---
title: "Introduction to the Multiplicative Half-Cauchy Process"
author: "Siddhesh Kulkarni"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the Multiplicative Half-Cauchy Process}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

library(mhcp)
```

# Introduction

The **Multiplicative Half-Cauchy Process (MHCP)** is an ordered
shrinkage process designed for settings in which progressively indexed
components should become smaller while retaining the possibility of
occasional large local excursions.

This type of behavior is useful in ordered latent representations, such
as factor models, where early components may capture substantial signal
while later components are increasingly shrunk toward zero.

The `mhcp` package provides computational tools for:

- simulating MHCP trajectories,
- evaluating component-wise probabilities,
- calibrating the global shrinkage parameter from interpretable
  probability statements,
- calculating local excursion probabilities, and
- summarizing the implied threshold-based effective dimension.

The package accompanies the manuscript:

> **Pathwise Shrinkage and Probabilistic Calibration of the
> Multiplicative Half-Cauchy Process**  
> Siddhesh Kulkarni (2026)

This vignette introduces the process and demonstrates the main package
functions.

# 1. Definition of the MHCP

Let

\[
\widetilde{\eta}_1 = 1
\]

and, for \(h \geq 2\),

\[
\widetilde{\eta}_h \sim C^+(0,\zeta),
\]

independently, where \(C^+(0,\zeta)\) denotes a half-Cauchy distribution
with scale parameter \(\zeta>0\).

The multiplicative process is

\[
\eta_h
=
\prod_{\ell=1}^{h}\widetilde{\eta}_\ell.
\]

Because \(\widetilde{\eta}_1=1\),

\[
\eta_1=1.
\]

For \(h\geq2\),

\[
\eta_h
=
\eta_{h-1}\widetilde{\eta}_h.
\]

Thus each new scale is obtained by multiplying the previous scale by an
independent half-Cauchy increment.

The parameter \(\zeta\) controls the long-run shrinkage behavior of the
process.

# 2. Simulation

The function `rmhcp()` simulates independent trajectories from the
process.

For example:

```{r simulate-basic}
set.seed(123)

eta <- rmhcp(
  n = 5,
  d = 10,
  zeta = 0.7
)

round(eta, 3)
```

Each row corresponds to an independent trajectory, while the columns
correspond to the ordered components

\[
\eta_1,\eta_2,\ldots,\eta_{10}.
\]

The first column is always equal to 1.

The process is not required to decrease at every step. Even when
\(\zeta<1\), individual half-Cauchy increments can exceed 1 and produce
temporary upward movements.

This distinction between **local behavior** and **long-run behavior** is
central to the MHCP.

# 3. Working on the log scale

Multiplicative processes are particularly convenient to study after
taking logarithms.

Write

\[
\widetilde{\eta}_h=\zeta U_h,
\]

where

\[
U_h\sim C^+(0,1).
\]

Define

\[
Y_h=\log U_h.
\]

Then

\[
\log\widetilde{\eta}_h
=
\log\zeta+Y_h,
\]

and therefore

\[
\log\eta_h
=
(h-1)\log\zeta
+
\sum_{\ell=2}^{h}Y_\ell.
\]

The multiplicative process has therefore become an additive stochastic
process on the log scale.

The package can simulate directly on this scale:

```{r simulate-log}
set.seed(123)

log_eta <- rmhcp(
  n = 5,
  d = 10,
  zeta = 0.7,
  log = TRUE
)

round(log_eta, 3)
```

Working on the log scale is also numerically useful for long
trajectories.

# 4. Distribution of the log increments

For a standard half-Cauchy random variable \(U\),

\[
f_U(u)
=
\frac{2}{\pi(1+u^2)},
\qquad u>0.
\]

With

\[
Y=\log U,
\]

the change of variables \(u=e^y\) gives

\[
f_Y(y)
=
\frac{1}{\pi\cosh(y)},
\qquad -\infty<y<\infty.
\]

This distribution is symmetric around zero.

Consequently,

\[
E(Y)=0
\]

and

\[
\operatorname{Var}(Y)
=
\frac{\pi^2}{4}.
\]

It follows that

\[
E(\log\eta_h)
=
(h-1)\log\zeta
\]

and

\[
\operatorname{Var}(\log\eta_h)
=
(h-1)\frac{\pi^2}{4}.
\]

The characteristic function of \(Y\) is

\[
\phi_Y(t)
=
\operatorname{sech}\left(\frac{\pi t}{2}\right).
\]

These properties provide the basis for the exact probability
calculations implemented in `mhcp`.

# 5. Median behavior

Because the log increments \(Y_h\) are symmetric around zero, the
centered sum

\[
\sum_{\ell=2}^{h}Y_\ell
\]

has median zero.

Therefore,

\[
\operatorname{Med}(\log\eta_h)
=
(h-1)\log\zeta,
\]

and hence

\[
\operatorname{Med}(\eta_h)
=
\zeta^{h-1}.
\]

Thus \(\zeta\) has a direct interpretation as the geometric rate of the
median trajectory.

For example, when \(\zeta=0.7\),

\[
\operatorname{Med}(\eta_h)
=
0.7^{h-1}.
\]

# 6. Pathwise shrinkage

The log-scale representation also reveals the long-run behavior of the
process.

Since the \(Y_h\)'s are independent and have finite mean zero, the
strong law of large numbers gives

\[
\frac{1}{h-1}
\sum_{\ell=2}^{h}Y_\ell
\longrightarrow
0
\qquad\text{almost surely}.
\]

Therefore,

\[
\frac{\log\eta_h}{h-1}
\longrightarrow
\log\zeta
\qquad\text{almost surely}.
\]

Equivalently,

\[
\eta_h^{1/(h-1)}
\longrightarrow
\zeta
\qquad\text{almost surely}.
\]

This produces three regimes.

If

\[
0<\zeta<1,
\]

then

\[
\eta_h\longrightarrow0
\qquad\text{almost surely}.
\]

If

\[
\zeta>1,
\]

then

\[
\eta_h\longrightarrow\infty
\qquad\text{almost surely}.
\]

The boundary case \(\zeta=1\) corresponds to a zero-drift random walk on
the log scale and does not produce systematic shrinkage.

For shrinkage applications, the primary regime of interest is therefore

\[
0<\zeta<1.
\]

# 7. Visualizing pathwise behavior

The following simulation illustrates the shrinking regime.

```{r pathwise-plot}
set.seed(123)

zeta <- 0.7
d <- 40

sim <- rmhcp(
  n = 25,
  d = d,
  zeta = zeta,
  log = TRUE
)

h <- seq_len(d)

matplot(
  h,
  t(sim),
  type = "l",
  lty = 1,
  xlab = "Component index h",
  ylab = "log(eta_h)"
)

lines(
  h,
  (h - 1) * log(zeta),
  lwd = 3,
  lty = 2
)

legend(
  "topright",
  legend = c(
    "Sample trajectories",
    "Theoretical log-median"
  ),
  lty = c(1, 2),
  lwd = c(1, 3),
  bty = "n"
)
```

Individual trajectories can fluctuate considerably. Nevertheless, their
long-run geometric rate is governed by \(\zeta\).

This is the sense in which the MHCP combines **local flexibility with
global shrinkage**.

# 8. Local multiplicative excursions

Because

\[
\frac{\eta_h}{\eta_{h-1}}
=
\widetilde{\eta}_h,
\]

the probability of a local multiplicative excursion has a simple closed
form.

For \(c>0\),

\[
P(\eta_h>c\eta_{h-1})
=
P(\widetilde{\eta}_h>c)
=
\frac{2}{\pi}
\arctan\left(\frac{\zeta}{c}\right).
\]

The package evaluates this probability using
`mhcp_excursion_prob()`.

For example:

```{r excursion-example}
mhcp_excursion_prob(
  zeta = 0.7,
  c = c(1, 2, 5)
)
```

For \(\zeta=0.7\), the probabilities are approximately:

```text
c = 1:  0.389
c = 2:  0.214
c = 5:  0.089
```

Thus, even though the process converges toward zero almost surely when
\(\zeta=0.7\), there remains substantial probability of local upward
movement.

Almost-sure shrinkage therefore should not be interpreted as
step-by-step monotonic shrinkage.

# 9. Distribution of a selected component

Define the centered log process

\[
Z_H
=
\sum_{\ell=2}^{H}Y_\ell.
\]

Then

\[
\log\eta_H
=
(H-1)\log\zeta+Z_H.
\]

Since the increments are independent,

\[
\phi_{Z_H}(t)
=
\left[
\operatorname{sech}
\left(
\frac{\pi t}{2}
\right)
\right]^{H-1}.
\]

The package evaluates the distribution of \(Z_H\) numerically using
characteristic-function inversion.

For example, symmetry implies

\[
P(Z_H\leq0)=0.5.
\]

We can verify this using:

```{r centered-cdf}
pmhcp_centered(
  q = 0,
  H = 10
)
```

The result is approximately `0.5`.

A selected quantile can be obtained with:

```{r centered-quantile}
qmhcp_centered(
  p = 0.90,
  H = 10
)
```

For \(H=10\), the 90th percentile is approximately

\[
q_{0.90,10}
\approx
5.967.
\]

# 10. Component probabilities

The function `pmhcp()` evaluates probabilities for the original
multiplicative process.

For example:

```{r component-cdf}
pmhcp(
  x = 0.05,
  H = 10,
  zeta = 0.369411
)
```

This is approximately `0.90`.

Thus a value near

\[
\zeta=0.3694
\]

implies that the 10th component is below \(0.05\) with approximately
90% prior probability.

# 11. Probability-based calibration

Rather than choosing \(\zeta\) through trial and error, the MHCP can be
calibrated from an interpretable prior statement.

Suppose an analyst specifies

\[
P(\eta_H\leq\varepsilon)=p.
\]

Let \(q_{p,H}\) denote the \(p\)-quantile of \(Z_H\).

Because

\[
\log\eta_H
=
(H-1)\log\zeta+Z_H,
\]

the calibration equation gives

\[
\zeta
=
\exp
\left\{
\frac{
\log\varepsilon-q_{p,H}
}{
H-1
}
\right\}.
\]

The function `calibrate_mhcp()` performs this calculation.

For example, suppose we want

\[
P(\eta_{10}\leq0.05)=0.90.
\]

Use:

```{r calibration-example}
cal90 <- calibrate_mhcp(
  H = 10,
  epsilon = 0.05,
  p = 0.90
)

cal90
```

The resulting value is approximately

\[
\zeta=0.3694.
\]

We can verify the result:

```{r calibration-verification}
pmhcp(
  x = 0.05,
  H = 10,
  zeta = cal90$zeta
)
```

which returns approximately `0.90`.

This is one of the main practical advantages of the MHCP formulation:
the hyperparameter can be connected directly to an interpretable prior
probability.

# 12. Median calibration

An especially simple case occurs when

\[
p=0.5.
\]

Because the centered distribution is symmetric,

\[
q_{0.5,H}=0.
\]

Therefore,

\[
\zeta
=
\varepsilon^{1/(H-1)}.
\]

For example:

```{r median-calibration}
cal50 <- calibrate_mhcp(
  H = 10,
  epsilon = 0.05,
  p = 0.50
)

cal50
```

Compare this with the closed-form calculation:

```{r median-check}
0.05^(1 / 9)
```

The two values agree.

# 13. Comparing calibration probabilities

Consider three different prior probability statements for the same
target component and threshold.

```{r calibration-comparison}
probabilities <- c(0.50, 0.75, 0.90)

calibrations <- do.call(
  rbind,
  lapply(
    probabilities,
    function(prob) {
      calibrate_mhcp(
        H = 10,
        epsilon = 0.05,
        p = prob
      )
    }
  )
)

calibrations
```

The calibrated values are approximately:

| Probability \(p\) | Calibrated \(\zeta\) |
|:-----------------:|:---------------------:|
| 0.50 | 0.717 |
| 0.75 | 0.508 |
| 0.90 | 0.369 |

As the desired probability increases, the calibrated value of \(\zeta\)
decreases.

This makes sense: stronger prior confidence that component \(H\) should
be negligible requires stronger shrinkage.

# 14. Normal approximation

The centered log process also satisfies a central limit theorem.

Because

\[
E(Y_h)=0
\]

and

\[
\operatorname{Var}(Y_h)=\frac{\pi^2}{4},
\]

we have

\[
\frac{
Z_H
}{
(\pi/2)\sqrt{H-1}
}
\Longrightarrow
N(0,1).
\]

This provides an approximate calibration method for sufficiently large
\(H\).

The package exposes this through:

```{r normal-calibration}
calibrate_mhcp(
  H = 10,
  epsilon = 0.05,
  p = 0.90,
  method = "normal"
)
```

The default is

```r
method = "exact"
```

which uses numerical inversion of the exact characteristic function.

For routine calibration, the exact method is recommended when
computationally convenient.

# 15. Fractional moments

The half-Cauchy distribution does not possess an ordinary first moment.

However, fractional moments exist.

For

\[
-1<q<1,
\]

a standard half-Cauchy random variable \(U\) satisfies

\[
E(U^q)
=
\sec\left(\frac{\pi q}{2}\right).
\]

Consequently,

\[
E(\eta_h^q)
=
\zeta^{q(h-1)}
\left[
\sec\left(\frac{\pi q}{2}\right)
\right]^{h-1}.
\]

Equivalently,

\[
E(\eta_h^q)
=
\left[
\zeta^q
\sec\left(\frac{\pi q}{2}\right)
\right]^{h-1}.
\]

These fractional moments are useful for establishing summability and
effective-dimension properties even though ordinary moments such as
\(E(\eta_h)\) are infinite.

This illustrates an important distinction: strong pathwise shrinkage
can coexist with heavy marginal tails.

# 16. Almost-sure summability

When \(0<\zeta<1\), the pathwise convergence is strong enough to imply
that, for every \(r>0\),

\[
\sum_{h=1}^{\infty}\eta_h^r
<
\infty
\qquad\text{almost surely}.
\]

Thus, although occasional large local excursions remain possible, a
typical infinite trajectory eventually shrinks rapidly enough to be
summable.

This provides a useful theoretical justification for interpreting the
MHCP as an ordered shrinkage mechanism.

# 17. Threshold-based effective dimension

For a practical threshold \(\varepsilon>0\), define

\[
N_\varepsilon
=
\#\{h:\eta_h>\varepsilon\}.
\]

This counts the number of components whose scales exceed a
user-specified threshold.

When \(0<\zeta<1\),

\[
N_\varepsilon<\infty
\qquad\text{almost surely}.
\]

For \(\varepsilon<1\),

\[
E(N_\varepsilon)
=
1+
\sum_{h=2}^{\infty}
P(\eta_h>\varepsilon).
\]

Equivalently,

\[
E(N_\varepsilon)
=
1+
\sum_{h=2}^{\infty}
\left[
1-
F_h
\left\{
\log\varepsilon-(h-1)\log\zeta
\right\}
\right],
\]

where \(F_h\) denotes the CDF of the centered log process at component
\(h\).

The package evaluates this quantity numerically using
`expected_mhcp_dimension()`.

# 18. Effective dimension after calibration

Consider again the calibration

\[
P(\eta_{10}\leq0.05)=0.90.
\]

We obtained approximately

\[
\zeta=0.3694.
\]

The corresponding expected effective dimension is:

```{r effective-dimension}
eff90 <- expected_mhcp_dimension(
  zeta = cal90$zeta,
  epsilon = 0.05
)

eff90$mean_effective_dimension
```

The result is approximately

\[
E(N_{0.05})=4.75.
\]

Thus the local calibration statement about component 10 implies a
global prior in which approximately 4.75 components, on average, remain
above the threshold \(0.05\).

# 19. Calibration and effective dimension together

The relationship can be illustrated for several probability levels.

```{r calibration-effective-dimension}
probabilities <- c(0.50, 0.75, 0.90)

results <- do.call(
  rbind,
  lapply(
    probabilities,
    function(prob) {

      cal <- calibrate_mhcp(
        H = 10,
        epsilon = 0.05,
        p = prob
      )

      eff <- expected_mhcp_dimension(
        zeta = cal$zeta,
        epsilon = 0.05
      )

      data.frame(
        p = prob,
        zeta = cal$zeta,
        expected_dimension = eff$mean_effective_dimension
      )
    }
  )
)

results
```

The approximate results are:

| \(p\) | \(\zeta\) | \(E(N_{0.05})\) |
|:---:|:---:|:---:|
| 0.50 | 0.717 | 20.63 |
| 0.75 | 0.508 | 7.60 |
| 0.90 | 0.369 | 4.75 |

Increasing the desired probability of negligible scale at component 10
therefore leads to both a smaller calibrated \(\zeta\) and a smaller
expected effective dimension.

# 20. A complete workflow

A typical MHCP analysis can be organized in four steps.

## Step 1: Specify an interpretable prior statement

Choose:

- a component \(H\),
- a practically negligible threshold \(\varepsilon\), and
- a desired probability \(p\).

For example:

```{r workflow-parameters}
H <- 10
epsilon <- 0.05
p <- 0.90
```

## Step 2: Calibrate the shrinkage parameter

```{r workflow-calibration}
cal <- calibrate_mhcp(
  H = H,
  epsilon = epsilon,
  p = p
)

cal
```

## Step 3: Examine the implied effective dimension

```{r workflow-dimension}
eff <- expected_mhcp_dimension(
  zeta = cal$zeta,
  epsilon = epsilon
)

eff$mean_effective_dimension
```

## Step 4: Simulate trajectories from the calibrated prior

```{r workflow-simulation}
set.seed(2026)

prior_paths <- rmhcp(
  n = 20,
  d = 30,
  zeta = cal$zeta,
  log = TRUE
)

matplot(
  seq_len(ncol(prior_paths)),
  t(prior_paths),
  type = "l",
  lty = 1,
  xlab = "Component index h",
  ylab = "log(eta_h)"
)
```

This workflow allows the analyst to examine the practical implications
of a probability-based calibration before fitting a larger Bayesian
model.

# 21. Main package functions

The primary user-facing functions are:

| Function | Description |
|:---|:---|
| `rmhcp()` | Simulate MHCP trajectories |
| `pmhcp()` | Evaluate \(P(\eta_H\leq x)\) |
| `pmhcp_centered()` | Evaluate the centered log-process CDF |
| `qmhcp_centered()` | Calculate centered log-process quantiles |
| `calibrate_mhcp()` | Calibrate \(\zeta\) from \(P(\eta_H\leq\varepsilon)=p\) |
| `mhcp_excursion_prob()` | Calculate local multiplicative excursion probabilities |
| `expected_mhcp_dimension()` | Calculate expected threshold-based effective dimension |

Help pages can be accessed using, for example:

```r
?rmhcp
?calibrate_mhcp
?expected_mhcp_dimension
```

# 22. Interpretation

The MHCP is designed to provide two properties that can otherwise be
difficult to combine.

First, it provides **global ordered shrinkage**. When \(0<\zeta<1\), the
process converges toward zero almost surely at a geometric rate governed
by \(\zeta\).

Second, it retains **local flexibility**. Individual multiplicative
increments can be large, allowing later components to temporarily
increase rather than enforcing deterministic or monotone shrinkage.

Probability-based calibration makes this behavior interpretable.
Instead of selecting \(\zeta\) as an abstract tuning parameter, an
analyst can specify a statement such as

\[
P(\eta_{10}\leq0.05)=0.90
\]

and calculate the corresponding prior.

The threshold-based effective dimension then provides a complementary
global summary of the resulting shrinkage behavior.

# 23. Reproducibility

The principal numerical example in the accompanying manuscript can be
reproduced with:

```{r final-reproducibility}
p_values <- c(0.50, 0.75, 0.90)

reproduction <- do.call(
  rbind,
  lapply(
    p_values,
    function(prob) {

      cal <- calibrate_mhcp(
        H = 10,
        epsilon = 0.05,
        p = prob
      )

      eff <- expected_mhcp_dimension(
        zeta = cal$zeta,
        epsilon = 0.05
      )

      data.frame(
        p = prob,
        q = cal$q,
        zeta = cal$zeta,
        expected_dimension = eff$mean_effective_dimension
      )
    }
  )
)

reproduction
```

The expected values are approximately:

| \(p\) | \(q_{p,10}\) | \(\zeta\) | \(E(N_{0.05})\) |
|:---:|:---:|:---:|:---:|
| 0.50 | 0.000 | 0.717 | 20.63 |
| 0.75 | 3.105 | 0.508 | 7.60 |
| 0.90 | 5.967 | 0.369 | 4.75 |

Small differences in the final displayed digits may occur because the
distribution and infinite-series calculations are evaluated
numerically.

# Citation

If you use `mhcp` in research, please cite:

> Kulkarni, S. (2026). *Pathwise Shrinkage and Probabilistic Calibration
> of the Multiplicative Half-Cauchy Process*. Manuscript in preparation.

Package citation information can also be obtained using:

```r
citation("mhcp")
```

The citation will be updated when final bibliographic information for
the accompanying article becomes available.
